pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools

### 2020-05-08 bmc_article.tex

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations (Figure \ref{fig:norm}B-C). Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations. In view of the subpopulation silhouette and ARI at true number of clusters, it seems that all methods perform worse on the three Zhengmix datasets, that are the datasets with the highest number of cells and the smaller feature detection rate. Scaling (e.g. `seurat (no scale)' vs `seurat') tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave equal  or poorer performance on MI, ARI and ARI at true number of clusters, with a notable negative impact when \textit{Seurat}'s regressing was applied to the Koh and Kumar datasets. It is possible that the difference between \textit{Seurat} and \textit{SCTransform}'s regression originates from the use of Pearson's residuals performed during the second round regression of \textit{SCTransform}, which are then independent of sequencing depth. \textit{SCTransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).\textit{scVI} (Supplementary Figure 15) performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy. For time and resource cost reasons, \textit{scVI} was run on a reduced set of pipelines and compared to \textit{SCTransform} (more details in the Methods section).

@@

We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:selection}B). To validate the previous assay on the proportion of variance/deviance explained by real populations, we included genes that maximized these two latter measures. Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets. The ARI at true number of clusters was near perfect (>0.94) among the smallest datasets but among bigger datasets (simMix and ZhengMix) the performance of the methods was lower and more variable (Figure \ref{fig:selection}B). Non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 22).

@@

The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Supplementary Figure 23). Again, the clustering on the biggest datasets (simMix and Zhengmix) was more challenging for most methods. Several of them could not identify the true number of clusters in the simMix1 and Zhengmix8eq datasets, which have the highest number of subpopulations, or lead to very low mean silhouette width and accuracy. Overall however, the clustering accuracy was higher for methods that include weighting or scaling. Among them, GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics. Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts. Its performance was lower than a pipeline with non-weighted \textit{Seurat} PCA on all measures for datasets with more subpopulations (Supplementary Figure 15). 
Overall, weighting the principal components by their variance (as \textit{Seurat} does) had a positive impact on silhouette widths and ARI scores.

@@

Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an 'elbow' (inflexion point) of the variance explained, to more complex algorithms. We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method (inflexion point in the variance explained by each component), some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.

We compared the various estimates of dimensionality in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space. As a first approximation of the true dimensionality, we computed the variance in each principal component that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:dim}A). Figure \ref{fig:dim}B shows the difference between the dimension estimates of the above methods and that based on the subpopulations (i.e. from Figure \ref{fig:dim}A). Of note, the methods differ widely in view of their computing time (Figure \ref{fig:dim}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method. Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:dim}C-D), when combined with \textit{SCTransform} or the standard \textit{Seurat} normalization. Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using the 20 nearest neighbors, although estimates were relatively stable across decent values of \textit{k}; see Supplementary Figure 26) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:dim}C) and resulted in the best clustering accuracy (Figure \ref{fig:dim}D). This method systematically selected many more components than were associated with the subpopulations, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification. An implication is that the commonly-used elbow method most likely underestimates dimensionality.

@@

@@

The basic pipeline presented here can be extended with additional analysis steps while keeping the same evaluation metrics. To demonstrate this, we evaluated various imputation or denoising techniques based on their impact on classification. Since preliminary analysis showed that all methods performed equally well or better on normalized data, we applied them after filtering and normalization, but before scaling and reduction. Although some of the methods (e.g., \textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Supplementary Figure 28A). When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Supplementary Figure 28B). However, the situation was not so straightforward with alternative metrics, where some methods (e.g., \textit{ENHANCE}) consistently underperformed. 10X datasets, which are typically characterized by a lower per-cell coverage and feature detection rate, benefited more from imputation and, in this context, \textit{DrImpute} and \textit{DCA} tended to show the best performance. On the contrary, imputing on normalized counts originating from Smart-seq technology was instead rather deleterious to the clustering accuracy.

@@

To illustrate the applicability of \textit{pipeComp} to a completely different setting, we benchmarked the impact of surrogate variable analysis (SVA) -based methods on bulk RNAseq differential expression analysis (DEA). The rationale behind such approaches is that the identification of factors determining genome-wide variation (i.e. surrogate variables) and their inclusion in the DEA's generalized linear model might improve DEA, especially in the presence of sources of variation unrelated to the experimental variables. \textit{sva}\citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq}\citep{rissoNormalizationRNAseq2014a} are two popular packages each implementing different methods to infer these variables. We therefore created a \texttt{PipelineDefinition} with three step: 1) feature filtering, 2) removal of unwanted variation (i.e. SVA), and 3) DEA using the most established methods. The package's `pipeComp\_dea' vignette details the creation process. We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citet{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A **simulation** of 8 vs 8 samples with two technical vectors of variation (see methods for more details). As previously reported, we found that more stringent filtering improved DEA (Supplementary Figure 29A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 29A). Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 29-30). With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C). While the FDR did increase in the \textit{ipsc} dataset, we suspect that much of the effect could be due to spurious differences \citep{germainTamingHumanGenetic2017} between the groups which we wrongly assume to be false positives, and the FDR nevertheless remained acceptable with 1-3 variables. These results therefore suggest that these methods are robust and can be widely applied.

@@

On the basis of our findings, we can make a number of practical analysis recommendations, also summarized in Supplementary Figure 31:

\begin{enumerate}
   \item Filtering
   \begin{itemize}
     \item Doublet detection and removal is advised and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions) to exclude poor-quality cells while avoiding bias against some subpopulations.
     \item Features filtering based on feature type did not appear beneficial.
   \end{itemize}
   \item Normalization and scaling
   \begin{itemize}
     \item Most normalization methods tested yielded a fair performance, especially when combined with scaling, which tended to have a positive impact on clustering.
     \item \textit{SCTransform} offered the best overall performance in terms of the separability of the subpopulations, as well as removing the effect of library size and detection rate.
     \item The common practice of regressing out cell covariates, such as the detection rate or proportion of mitochondrial reads nearly always had a negative impact, leading to increased correlation with covariates and decreased clustering accuracy. We therefore advise against this practice.
   \end{itemize}
   \item Feature selection
   \begin{itemize}
     \item Deviance \cite{townesGlmpca2019} offered the best ranking of genes for feature selection.
     \item Increasing the number of features included tended to lead to better classifications, plateauing from 4000 features in our datasets.
   \end{itemize}
   \item Denoising/imputation
   \begin{itemize}
       \item Denoising appeared beneficial to the identification of subpopulations in 10x datasets, but not in Smart-seq datasets.
       \item We found especially \textit{ALRA} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{DCA} to offer the best performances.
   \end{itemize}
   \item PCA
   \begin{itemize}
     \item Similarly to previous reports \cite{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \textit{Seurat}. 
     \item We advise against the commonly-used elbow method (because it is too conservative) and the jackstraw method (low performance at great computational costs) for deciding on the components to include, and recommend a method like the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g., implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}, in our case using the 20 nearest neighbors).
   \end{itemize}
   \item Clustering
   \begin{itemize}
    \item We found \textit{scran}-based \textit{walktrap} clustering to show good performances, at however increased computational costs which might be prohibitive with large datasets.
    \item In cases where prior knowledge can guide the choice of a resolution, \textit{Seurat} can be useful in affording manual control of it while, in the absence of such knowledge, \textit{scran}-based \textit{walktrap} clustering provided reasonable estimates.
   \end{itemize}
\end{enumerate}

@@

Among all pipelines that were tested, the performance varied greatly among datasets. The ZhengMix8eq and simMix1 datasets were the most challenging datasets across all preprocessing steps that we tested. The average silhouette width of half of their subpopulations was systematically lower that the other subpopulations/ datasets and the ARI score at true number of clusters was always poor or unable to achieve. In comparison to the performance on other datasets, this inconsistency was particularly striking when evaluating the dimensionality estimation dimensionality reduction and the clustering methods, for which half of the evaluated methods could not lead to the true number of clusters. It is likely that these two datasets are challenging due to their combined sparsity and complexity (see Supplementary Figure 1). This issue that was already observed for the Zhengmix8eq dataset in a previous benchmark \cite{duoClustering2018}, as it contains deeper levels of annotations for subpopulations of T-cells. The Koh dataset harbors as many populations, but originates from more distinct groups (from hESC in different differentiation stages). As the Kumar dataset, it harbors a very low subpopulations complexity and most methods applied on them performed well.

@@

We developed \textit{pipeComp} to address the need of a framework that can simultaneously evaluate the interaction of multiple tools. The work of Vieth and colleagues\cite{viethSystematic2019} already offered an important precedent in this respect, evaluating the interaction of various steps in the context of scRNAseq differential expression analysis, but did not offer a platform for doing so. Instead, the \textit{CellBench} package was recently proposed to address a similar need \cite{suCellbench}, offering an elegant piping syntax to combine alternative methods at successive steps. A key additional feature of \textit{pipeComp} is its ability to perform evaluation in parallel to the pipeline and offering on-the-fly evaluation at any step. This avoids the need to store potentially large intermediate data for all possible permutations and thus allowing a combinatorial benchmark. In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, making it easy to modify, extend with further methods, or apply to other datasets.

@@

\textit{pipeComp} is a flexible R framework for evaluating methods and parameter combinations in a complex pipeline, computing on-the-fly multilevel evaluation metrics. Applying this framework to scRNAseq clustering enabled us to make clear recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Supplementary Figure 31). We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance. We hope that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to investigate pipelines other domains with multi-step computational analyses.

@@

@@

@@

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/dimensionality.pdf}}
    \fi
    \caption{\textbf{Estimating dimensionality. A:} Estimated dimensionality using the proportion of variance in each component explained by the true subpopulations. \textbf{B:} Difference between `real' dimensionality (from \textbf{A}) and dimensionality estimation methods, along with computing time. \textbf{C:} Average silhouette width per subpopulation across a selection of methods, combined with \textit{SCTransform} or standard \textit{Seurat} normalization. \textbf{D:} Clustering accuracy across the normalization/dimensionality estimation methods. The color-mapping scheme is the same as in Figure \ref{fig:norm}}
    \label{fig:dim}
\end{figure}

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/clustering.pdf}}
    \fi
    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{SCTransform} and standard \textit{Seurat}'s normalization.  The color-mapping scheme is the same as in Figure \ref{fig:norm}}
    \label{fig:clust}
\end{figure}

>>>>

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations. Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI, Figure \ref{fig:norm}B-C). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations, and that the same datasets (e.g. Zhengmix8eq and simMix1) and subpopulations were most challenging across for all methods. An exception was \textit{scVI}, which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy (Supplementary Figure 13; \textit{scVI} was run on a reduced set of pipelines). Scaling (e.g. `seurat (no scale)' vs `seurat') tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave equal  or poorer performance on MI, ARI and ARI at true number of clusters, with a notable negative impact in the Smart-seq datasets (Koh and Kumar). When performed with \textit{SCTransform}, regressing out covariates using tended to have neither a positive or negative effect in droplet-based datsets, possibly owing to its regularization.  In general, \textit{SCTransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).

@@

We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:selection}B). To validate the previous assay on the proportion of variance/deviance explained by real populations, we included genes that maximized these two latter measures. Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets (Figure \ref{fig:selection}B). Non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also tended to lead to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 22).

@@

The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Supplementary Figure 23). Overall, weighting the components by the variance their explain tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary Figure 23B). GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics. Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts. Its performance was however the worst both in terms of silhouette width and clustering accuracy (Supplementary Figure 15), and it only performed well in some of the least challenging datasets (mixology10x5cl and simMix2).

@@

Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an inflexion point in the curve of the variance explained by each component (`elbow' method), to more complex algorithms. We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method, some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.

We compared the various estimates of in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space. As a first approximation of the true dimensionality, we computed the variance in each principal component (based on Seurat's weighted PCA space) that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:dim}A). These truth-based estimates where then compared to those of the above dimensionality estimation methods (Figure \ref{fig:dim}B). Of note, the methods differ widely in view of their computing time (Figure \ref{fig:dim}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method. Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:dim}C-D), when combined with \textit{SCTransform} or standard \textit{Seurat} normalization. Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using 10 or 20 nearest neighbors, although estimates were relatively stable across decent values of \textit{k}; see Supplementary Figure 26) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D). This method systematically selected many more components than were associated with the subpopulations in Figure \ref{fig:dim}A, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification. An important implication is that the commonly-used elbow method most likely underestimates dimensionality.

@@

\subsection*{Overview of interactions between steps}

We selected the most important alternatives for each parameter (see methods) in order to explore their interactions. Analysis of variance (after correcting for baseline differences between datasets and the resolution parameter) attributed the largest proportion of variance in the main cluster metrics to the choice of dimensionality, followed by normalization, and very little to filtering (Supplementary Figure 28). Because the choice of dimensions impacts the number of clusters, we repeated this analysis of variance among clusterings that yielded the true number of clusters, and observed that although normalization was the most important parameter among these analyses, the choice of dimensionality came closely behind (Supplementary Figure 28B), suggesting that more attention should be paid to this relatively neglected step.

We next investigated whether the benefits of some methods where conditional on choices at other steps. Although analysis of variance identified many weak but significant interactions, in general we found that the benefits brought by important methods were robust to differences in other parts of the pipeline (Supplementary Figures 29-30). For instance, we confirmed that doublet removal and \textit{SCtransform} normalization were beneficial independently of other steps, and that, again indepdendently of other steps, more stringent filtering did not yield any major improvement over lenient filtering (Supplementary Figure 29). Similarly, deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Supplementary Figure 30).

@@

The basic pipeline presented here can be extended with additional analysis steps while keeping the same evaluation metrics. To demonstrate this, we evaluated various imputation or denoising techniques based on their impact on classification. Since preliminary analysis showed that all methods performed equally well or better on normalized data, we applied them after filtering and normalization, but before scaling and reduction. Although some of the methods (e.g., \textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Supplementary Figure 31A). When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Supplementary Figure 31B). However, the situation was not so straightforward with alternative metrics, where some methods (e.g., \textit{ENHANCE}) consistently underperformed. 10X datasets, which are typically characterized by a lower per-cell coverage and feature detection rate, benefited more from imputation and, in this context, \textit{DrImpute} and \textit{DCA} tended to show the best performance. On the contrary, imputing on normalized counts originating from Smart-seq technology was instead rather deleterious to the clustering accuracy.

@@

To illustrate the applicability of \textit{pipeComp} to a completely different problem, we benchmarked the impact of surrogate variable analysis (SVA) -based methods on bulk RNAseq differential expression analysis (DEA). The rationale behind such approaches is that the identification of factors determining genome-wide variation (i.e. surrogate variables) and their inclusion in the DEA's generalized linear model might improve DEA, especially in the presence of sources of variation unrelated to the experimental variables. \textit{sva}\citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq}\citep{rissoNormalizationRNAseq2014a} are two popular packages each implementing different methods to infer these variables. We therefore created a \texttt{PipelineDefinition} with three step: 1) feature filtering, 2) removal of unwanted variation (i.e. SVA), and 3) DEA using the most established methods. The package's `pipeComp\_dea' vignette details the creation process. We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citet{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A \texttt{simulation} of 8 vs 8 samples with two technical vectors of variation (see methods for more details). As previously reported, we found that more stringent filtering improved DEA (Supplementary Figure 32A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 32A). Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 32-33). With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C). While the FDR did increase in the \textit{ipsc} dataset, we suspect that much of the effect could be due to spurious differences \citep{germainTamingHumanGenetic2017} between the groups which we wrongly assume to be false positives, and the FDR nevertheless remained acceptable with 1-3 variables. These results therefore suggest that these methods are robust and can be widely applied.

@@

On the basis of our findings, we can make a number of practical analysis recommendations, also summarized in Supplementary Figure 34:

\begin{enumerate}
   \item Filtering
   \begin{itemize}
     \item Doublet detection and removal is advised and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions) to exclude poor-quality cells while avoiding bias against some subpopulations.
     \item Features filtering based on feature type did not appear beneficial.
   \end{itemize}
   \item Normalization and scaling
   \begin{itemize}
     \item Most normalization methods tested yielded a fair performance, especially when combined with scaling, which tended to have a positive impact on clustering.
     \item \textit{SCTransform} offered the best overall performance in terms of the separability of the subpopulations, as well as removing the effect of library size and detection rate.
     \item The common practice of regressing out cell covariates, such as the detection rate or proportion of mitochondrial reads nearly always had a negative impact, leading to increased correlation with covariates and decreased clustering accuracy. We therefore advise against this practice.
   \end{itemize}
   \item Feature selection
   \begin{itemize}
     \item Deviance \cite{townesGlmpca2019} offered the best ranking of genes for feature selection.
     \item Increasing the number of features included tended to lead to better classifications, plateauing from 4000 features in our datasets.
   \end{itemize}
   \item Denoising/imputation
   \begin{itemize}
       \item Denoising appeared beneficial to the identification of some subpopulations in 10x datasets, but not in Smart-seq datasets.
       \item We found especially \textit{ALRA} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{DCA} to offer the best performances, although none of these methods was beneficial over all subpopulations.
   \end{itemize}
   \item PCA
   \begin{itemize}
     \item Similarly to previous reports \cite{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \textit{Seurat}. 
     \item We advise against the commonly-used elbow method (because it is too conservative) and the jackstraw method (low performance at great computational costs) for deciding on the components to include, and recommend a method like the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g., implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}, in our case using the 10 or 20 nearest neighbors).
   \end{itemize}
   \item Clustering
   \begin{itemize}
    \item We found \textit{scran}-based \textit{walktrap} clustering to show good performances, at however increased computational costs which might be prohibitive with large datasets.
    \item In cases where prior knowledge can guide the choice of a resolution, \textit{Seurat} can be useful in affording manual control of it while, in the absence of such knowledge, \textit{scran}-based \textit{walktrap} clustering provided reasonable estimates.
   \end{itemize}
\end{enumerate}

@@

Among all pipelines that were tested, the performance varied greatly among datasets. The ZhengMix8eq and simMix1 datasets were the most challenging datasets across all preprocessing steps that we tested. The average silhouette width of half of their subpopulations was systematically lower that the other subpopulations/ datasets and the ARI score at true number of clusters, if ever achieved, was always poor. It is likely that these two datasets are challenging due to their combined sparsity and complexity (see Supplementary Figure 1). The challenge of distinguishing subpopulations of T-cells in the Zhengmix8eq dataset was previously discussed \cite{duoClustering2018}. In contrast, the Koh dataset harbors as many populations, but originates from more distinct groups (from hESC in different differentiation stages), while the Kumar and mixology datasets harbor a very low subpopulations complexity and most methods applied on them performed well.

@@

We developed \textit{pipeComp} to address the need of a framework that can simultaneously evaluate the interaction of multiple tools. The work of Vieth and colleagues\cite{viethSystematic2019} already offered an important precedent in this respect, evaluating the interaction of various steps in the context of scRNAseq differential expression analysis, but did not offer a platform for doing so. Instead, the \textit{CellBench} package was recently proposed to address a similar need \cite{suCellbench}, offering an elegant piping syntax to combine alternative methods at successive steps. A key additional feature of \textit{pipeComp} is its ability to perform evaluation in parallel to the pipeline and offering on-the-fly evaluation at multiple steps. This avoids the need to store potentially large intermediate data for all possible permutations and thus allowing a combinatorial benchmark. This, along with the set-up simplicity (demonstrated here by our separate SVA-DEA benchmark), also distinguishes \textit{pipeComp} from general workflow frameworks such as \textit{drake}\citep{landauDrakePackage2018}; indeed we estimate that the benchmarks discussed here would have taken approximately 16TB of storage had all non-redundant intermediates been saved (see the Supplementary methods for more detailed explanations of the framework and comparison to \textit{CellBench} and \textit{drake}). In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, making it easy to modify, extend with further methods, or apply to other datasets.

@@

\textit{pipeComp} is a flexible R framework for evaluating methods and parameter combinations in a complex pipeline, computing on-the-fly multilevel evaluation metrics. Applying this framework to scRNAseq clustering enabled us to make clear recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Supplementary Figure 34). We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance. We further showed that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to investigate pipelines other domains with multi-step computational analyses.

@@

\subsection*{Overall run for interactions across steps}

The set of alternatives found for the overall run (Supplementary Figures 28-30) can be found in the \href{https://github.com/markrobinsonuzh/scRNA_pipelines_paper/blob/master/data/overall_run.Rmd}{paper's repository}. The analysis of variance was performed with the following model:

\texttt{metric$\sim$ dataset*resolution+doubletmethod+filt+norm+clustmethod+dims}

@@

For \textit{vstsva}, we first applied the variance-stabilizing transformation from \textit{DESeq2}\citep{loveModeratedEstimation2014} before applying \texttt{sva::sva}. \textit{edgeR.QLF} refers to the quasi-likelihood version \citep{lundDetectingDifferential2012} of \textit{edgeR}'s fit. The exact wrappers used can be found in the package at \url{https://github.com/plger/pipeComp/blob/master/inst/extdata/dea\_wrappers.R}.

@@

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/dimensionality.pdf}}
    \fi
    \caption{\textbf{Estimating dimensionality. A:} Estimated dimensionality using the proportion of variance in each component explained by the true subpopulations. \textbf{B:} Difference between `real' dimensionality (from \textbf{A}) and dimensionality estimation methods, along with computing time. \textbf{C:} Average silhouette width per subpopulation across a selection of methods, combined with \textit{SCTransform} or standard \textit{Seurat} normalization. \textbf{D:} Clustering accuracy across the normalization/dimensionality estimation methods. The color-mapping scheme is the same as in Figure \ref{fig:norm}C}
    \label{fig:dim}
\end{figure}

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/clustering.pdf}}
    \fi
    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{SCTransform} and standard \textit{Seurat}'s normalization.  The color-mapping scheme in panel \textbf{A} and the left of panel \textbf{B} is the same as in Figure \ref{fig:norm}C.}
    \label{fig:clust}
\end{figure}
